Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2021-10-06 methods_paper/fabl-paper.tex
update.
added demonstration, addressed Bekas points

The likelihood becomes 
\begin{align}
\mathcal{L}(\bm{Z}, \Phi \mid \bm{\gamma}^{obs}) = 
\prod_{j=1}^{n_1}  \prod_{i=1}^{n_2}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{a_{f\ell}(\bm{Z})}u_{fl}^{a_{f\ell}(\bm{Z})}\right], 
\end{align}
where 
\begin{align}
&a_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f)I(\gamma_{ij}^f = \ell) I(Z_j = i) \\
&b_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f) I(\gamma_{ij}^f = \ell) I(Z_j \neq i),
\end{align}
where $I_{obs}$ denotes if a value is observed or not. Next, for a particular matching configuration $\bm{Z}$, $a_{f\ell}(\bm{Z})$
and $b_{f\ell}(\bm{Z})$ denote the number of matches and non-matches with observed disagreement level $\ell$ for comparison feature $f.$ Using this parameterization, we specify independent conjugate priors for all features $f$:
\begin{align}
&\bm{m}_f \sim \text{Dirichlet}(\alpha_{f0}, \ldots, \alpha_{f L_f}) \\
&\bm{u}_f \sim \text{Dirichlet}(\beta_{f0}, \ldots, \beta_{f L_f}).
\end{align}

@@

\begin{align*}
p(\mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda|\Gamma) &\propto p(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}) p(\mathbf{Z} | \lambda) p(\mathbf{m}, \mathbf{u}) p(\lambda) \\
&\propto \prod_{j=1}^{n_B}  \prod_{i=1}^{n_A}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right] \\
&\times  \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{\alpha_{fl} - 1}  \times\prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}^{\beta_{fl} - 1} \\
&\times \prod_{j=1}^{n_B} \left[I(Z_j \leq n_A)\frac{1}{n_A}\lambda + I(Z_j = n_A + 1)(1 - \lambda)\right]
\end{align*}
This factorization leads to following Gibbs Sampler:

@@

\underline{Sample $\lambda^{(s+1)}|\mathbf{Z}^{(s)}$:} As a function of
\(\lambda\), the linkage structure parameter \(\mathbf{Z}\) is sequence
of successes (when \(z_j < n_A + 1\)) and failures (when
\(z_j = n_A + 1\)), and therefore
\(p(\mathbf{Z}|\lambda) = \mathcal{L}(\lambda|\mathbf{Z})\) is
determined only by the number of duplicates
\(D = \sum_{i=1}^{n_B}\mathbf{1}_{z_j < n_A + 1}\) encoded by
\(\mathbf{Z}\). Thus we have

\[p(\lambda | \mathbf{Z}) \propto p(\mathbf{Z}|\lambda)p(\lambda)\]
\[\propto \lambda^D (1-\lambda)^{n_B - D} \lambda^{\alpha_{\lambda} -1} (1-\lambda)^{\beta_{\lambda} -1}\]
\[ \propto \lambda^{D + \alpha_{\lambda} - 1} (1-\lambda)^{n_B - D + \beta_{\lambda} -1}\]
\[\implies \lambda^{(s+1)}|\mathbf{Z}^{(s+1)} \sim \text{Beta}(D + \alpha_{\lambda}, n_B - D + \beta_{\lambda})\]

@@

\begin{align*}
	p(\Gamma_{.j}|Z_j = z_j, \mathbf{m}, \mathbf{u}) &\propto \prod_{i=1}^{n_A}\left[\prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)}\\
	&\propto \prod_{i=1}^{n_A}\left(\prod_{f=1}^{F}\prod_{l=1}^{L_f} \frac{m_{fl}}{u_{fl}}\right)^{I(z_j = i, \gamma_{ij}^f = l)} \\
	&=
	\begin{cases} 
		w_{ij}  & z_j \leq n_A; \\
		1 &  z_j  = n_A + 1 \\
	\end{cases}\\
\end{align*}

@@

We calculate a Bayes estimate \(\hat{\mathbf{Z}}\) for the linkage
parameter \(\mathbf{Z}\) by assigning different positive losses to
different types of errors, and minimizing posterior expected loss. We
adopt the loss function proposed in Sadinle 2017, in which
\(\hat{Z}_j \in \{1, \ldots, n_A + 1, R\}\), with \(R\) representing the
option to leave the matching undetermined by the model. Specifically, we
minimize the quantity
\(L(\hat{\mathbf{Z}}, \mathbf{Z}) = \sum_{j=1}^{n_B} L(\hat{Z_j}, Z_j)\)
where

\[L(\hat{Z_j}, Z_j)=\begin{cases} 
	0  & \text{if } Z_j = \hat{Z_j}; \\
	\theta_R,  & \text{if } \hat{Z_j} = R; \\
	\theta_{10},  & \text{if } Z_j \leq n_A,\hat{Z_j} = n_A + 1 ; \\
	\theta_{01},  & \text{if } Z_j = n_A + 1,\hat{Z_j} \leq n_A ; \\
	\theta_{11},  & \text{if } Z_j, \hat{Z}_j \leq n_A, Z_j \neq \hat{Z_j} ; \\
\end{cases}\] Here, \(\theta_R\) is the loss from not making a decision
on the linkage status, \(\theta_{10}\) is the loss from a false
non-match, \(\theta_{01}\) is the loss from a false match, and
\(\theta_{11}\) is the loss from the special case of a false match in
which the record has a true match other than the one estimated by the
model. This loss function leads to closed form decision rules for
minimizing posterior expected loss. In this paper, we adopt losses
\(\theta_R = \infty, \theta_{10} = 1, \theta_{01} = 1, \theta_{11} = 2\),
inducing the intuitive decision rule

@@

Since our Gibbs procedure does not strictly enforce one-to-one matching,
it is possible for the final Bayes estimate to link multiple records in
\(B\) to one record in \(A\). The modeler can either report both such
matches (with their respective posterior match probabilities), or
resolve these conflicts by accepting only the match with highest
posterior probability. A similar approach can be see in the most probable maximal matching sets used by \citep{steorts_2013} to match records to latent entities. Such a resolution procedure indeed is equivalent
to minimizing posterior risk under the restriction of one-to-one
matching, and as thus theoretically justified within the Bayesian
framework.

\section{Efficient and Scalable Implementation}
\label{sec:efficiency}
In this section, we propose efficient and scalable proposals using hashing based methods that are similar in spirit to \citep{enamorado2019}, a fast and scalable implementation of the Fellegi-Sunter model. 
 Specifically, one way we can improve our computational efficiency is by recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. To make this more precise, let 
\(\mathcal{H}\) be the set of unique agreement patterns in the data, let
\(P = |\mathcal{H}|\) denote the total number of unique agreement
patterns.  Observe that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and
that this bound \textcolor{red}{does not depend on} \(n_1\) or \(n_2\). Prior to
processing the data, we identify all \(P\) patterns in \(\mathcal{H}\)
and enumerate them as follows: \(h_1, \ldots, h_P\), which are called \emph{hashed values.}
Next, we map record pairs to their corresponding hashed value. 
More specifically, when the record pair \((i,j)\) 
exhibits the \(p^{th}\) agreement pattern, we say \((i,j) \in h_p\).
Whenever possible, we conduct calculations over these \(P\) agreement
patterns, instead of the typical  \(n_1 \times n_2\) record pairs.

@@

\textcolor{red}{RCS: I stopped here.}

@@

@@

@@

@@

@@

>>>>

The likelihood becomes 
\begin{align}
\mathcal{L}(\bm{Z}, \Phi \mid \bm{\Gamma}^{obs}) = 
\prod_{j=1}^{n_1}  \prod_{i=1}^{n_2}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{a_{f\ell}(\bm{Z})}u_{fl}^{a_{f\ell}(\bm{Z})}\right], 
\end{align}
where 
\begin{align}
&a_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f)I(\gamma_{ij}^f = \ell) I(Z_j = i) \\
&b_{f\ell}(\bm{Z}) = \sum_{i,j} I_{obs}(\gamma_{ij}^f) I(\gamma_{ij}^f = \ell) I(Z_j \neq i),
\end{align}
where $I_{obs}$ denotes if a value is observed or not. Next, for a particular matching configuration $\bm{Z}$, $a_{f\ell}(\bm{Z})$
and $b_{f\ell}(\bm{Z})$ denote the number of matches and non-matches with observed disagreement level $\ell$ for comparison feature $f.$ Using this parameterization, we specify independent conjugate priors for all features $f$:
\begin{align}
&\bm{m}_f \sim \text{Dirichlet}(\alpha_{f0}, \ldots, \alpha_{f L_f}) \\
&\bm{u}_f \sim \text{Dirichlet}(\beta_{f0}, \ldots, \beta_{f L_f}).
\end{align}

@@

\begin{align*}
p(\mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda|\Gamma) &\propto \mathcal{L}(\bm{Z}, \Phi \mid \bm{\gamma}^{obs}) p(\mathbf{Z} | \lambda) p(\Phi) p(\lambda) \\
&\propto \prod_{j=1}^{n_2}  \prod_{i=1}^{n_1}\left[ \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)} \\
&\times  \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{\alpha_{fl} - 1}  \times\prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}^{\beta_{fl} - 1} \\
&\times \prod_{j=1}^{n_B} \left[I(Z_j \leq n_1)\frac{1}{n_1}\lambda + I(Z_j > n_1)(1 - \lambda)\right]
\end{align*}
This factorization leads to following Gibbs Sampler:

@@

\underline{Sample $\lambda^{(s+1)}|\mathbf{Z}^{(s)}$:} As a function of
\(\lambda\), the linkage structure parameter \(\mathbf{Z}\) is sequence
of successes (when \(z_j < n_A + 1\)) and failures (when
\(z_j = n_A + 1\)), and therefore
\(p(\mathbf{Z}|\lambda) = \mathcal{L}(\lambda|\mathbf{Z})\) is
determined only by the number of duplicates
$n_{12}(\mathbf{Z}) = \sum_{i=1}^{n_2}I(z_j < n_1 + 1)$ encoded by
$\mathbf{Z}$. Thus we have

\[p(\lambda | \mathbf{Z}) \propto p(\mathbf{Z}|\lambda)p(\lambda)\]
\[\propto \lambda^{n_{12}(\mathbf{Z})} (1-\lambda)^{n_2 - (n_{12}(\mathbf{Z})} \lambda^{\alpha_{\lambda} -1} (1-\lambda)^{\beta_{\lambda} -1}\]
\[ \propto \lambda^{n_{12}(\mathbf{Z}) + \alpha_{\lambda} - 1} (1-\lambda)^{n_1 - n_{12}(\mathbf{Z}) + \beta_{\lambda} -1}\]
\[\implies \lambda^{(s+1)}|\mathbf{Z}^{(s+1)} \sim \text{Beta}(D + \alpha_{\lambda}, n_2 - n_{12}(\mathbf{Z}) + \beta_{\lambda})\]

@@

\begin{align*}
	p(\Gamma_{.j}|Z_j = z_j, \mathbf{m}, \mathbf{u}) &\propto \prod_{i=1}^{n_1}\left[\prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(Z_j = i)}u_{fl}^{I(Z_j \neq i)}\right]^{I(\gamma_{ij}^f = l)}\\
	&\propto \prod_{i=1}^{n_1}\left(\prod_{f=1}^{F}\prod_{l=1}^{L_f} \frac{m_{fl}}{u_{fl}}\right)^{I(z_j = i, \gamma_{ij}^f = l)} \\
	&=
	\begin{cases} 
		w_{ij}  & z_j \leq n_1; \\
		1 &  z_j  = n_1 + j \\
	\end{cases}\\
\end{align*}

@@

We calculate a Bayes estimate \(\hat{\mathbf{Z}}\) for the linkage
parameter \(\mathbf{Z}\) by assigning different positive losses to
different types of errors, and minimizing posterior expected loss. We
adopt the loss function proposed in Sadinle 2017, in which
\(\hat{Z}_j \in \{1, \ldots, n_A + 1, R\}\), with \(R\) representing the
option to leave the matching undetermined by the model. Specifically, we have

\[L(\hat{Z_j}, Z_j)=\begin{cases} 
	0  & \text{if } Z_j = \hat{Z_j}; \\
	\theta_R,  & \text{if } \hat{Z_j} = R; \\
	\theta_{10},  & \text{if } Z_j \leq n_A,\hat{Z_j} = n_A + 1 ; \\
	\theta_{01},  & \text{if } Z_j = n_A + 1,\hat{Z_j} \leq n_A ; \\
	\theta_{11},  & \text{if } Z_j, \hat{Z}_j \leq n_A, Z_j \neq \hat{Z_j} ; \\
\end{cases}\] Here, \(\theta_R\) is the loss from not making a decision
on the linkage status, \(\theta_{10}\) is the loss from a false
non-match, \(\theta_{01}\) is the loss from a false match, and
\(\theta_{11}\) is the loss from the special case of a false match in
which the record has a true match other than the one estimated by the
model.

We seek to minimize the posterior expected loss, given by $\mathbb{E}[L(\hat{\mathbf{Z}}, \mathbf{Z})|\Gamma^{obs}] = \sum_{\mathbf{Z}} L(\hat{Z_j}, Z_j)P(\mathbf{Z}|\Gamma^{obs})$. In this paper, we adopt losses
\(\theta_R = \infty, \theta_{10} = 1, \theta_{01} = 1, \theta_{11} = 2\), inducing the intuitive decision rule

@@

{
\color{blue}
Since our Gibbs procedure does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $\bm{X_2}$ to one record in $\bm{X_1}$. It is often desirable to see such matches, because they provide evidence of model misspecification; specifically, large numbers of "many to one" matches indicate that the assumption of no duplicates within files may not be reasonable. We will explore this issue more deeply through the El Salvador homicide case study.

To achieve a Bayes estimate that fulfills one-to-one matching requirement, we simply minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. In the event that two matches. In the event that we have two records $j$ and $j'$ such that both $P(Z_j = i |\Gamma) > \frac{1}{2}$ and $ P(Z_{j'} = i |\Gamma) > \frac{1}{2}$, this constraint means we accept the match with the highest posterior probability, and declare the other to have no match.  A similar approach can be see in the most probable maximal matching sets used by \citep{steorts_2013} to match records to latent entities, and is justified within this Bayesian framework.
}

\section{Efficient and Scalable Implementation}
\label{sec:efficiency}
In this section, we propose efficient and scalable proposals using hashing based methods that are similar in spirit to \citep{enamorado2019}, a fast and scalable implementation of the Fellegi-Sunter model. 
 Specifically, one way we can improve our computational efficiency is by recognizing that record pairs contribute to posterior calculations only
through the agreement pattern of the \(\gamma_{ij}\) vector. To make this more precise, let 
\(\mathcal{H}\) be the set of unique agreement patterns in the data, let
\(P = |\mathcal{H}|\) denote the total number of unique agreement
patterns.  Observe that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and
that this bound does not depend on \(n_1\) or \(n_2\). Prior to
processing the data, we identify all \(P\) patterns in \(\mathcal{H}\)
and enumerate them as follows: \(h_1, \ldots, h_P\), which are called \emph{hashed values.}
Next, we map record pairs to their corresponding hashed value. 
More specifically, when the record pair \((i,j)\) 
exhibits the \(p^{th}\) agreement pattern, we say \((i,j) \in h_p\).
Whenever possible, we conduct calculations over these \(P\) agreement
patterns, instead of the typical  \(n_1 \times n_2\) record pairs.

@@

@@

\textcolor{blue}{For Beka: The simulated files contain 6 fields of information, but he only uses 4 in his study. One of the fields was giving me problems, so I had been using a different one instead. I just figured out how to make the field work, so I can rerun the code. At that point, it will be a 100\% true replication.}

@@

{
\color{blue}For Beka: Yes, I used Table 2 (those are default settings in BRL)

My Figure 1 is equivalent to Sadinle's figure 3. He compares against Fellegi Sunter, whereas I compare against BRL. I found his style of graphic difficult to read, especcially because I wanted to show F-measure in addition to recall and precision.

Sadinle's Figure 4 just shows the results when you add the option of not declaring matching status (leaving it undecided by the model). This didn't seem to be so relevant in my paper. I can make the plot by changing just a few lines of code, so I can do it if you think I need it!}

@@

{
\color{blue} For Beka: I'm not using any blocking for either method.

Some of the hashing techniques could work for BRL (updating the m and u parameters), and they technically could be used for updating Z, but it would be much less useful because BRL requires knowing which exact record you match to, not just the agreement pattern for that record. Its not feasible for me try making those changes since its coded in C.

I'll work out the computational complexity. That can be one of the things we nail down on Thursday!}

@@

\textcolor{blue}{This is with post processing already. All I meant here was that is that small tasks require less information. If you do a $500 \times 500$ linkage task and you only have first and last name (for example), thats likely enough to get all the correct matches. However, once you do $20000 \times 20000$, its more likely that more than one person in the second file has the same first and last name, so you need more information to distinguish. I mentioned before that the simulated files actually have 6 fields of information, but I only used 4 originally because thats what Sadinle did. I can use all 6 though to see what happens?}

@@

 \textcolor{blue}{When I use the compareRecords from BRL to do the pairwise comparison on the concatenated files, its too  big for R to handle! One could do hashing, reduce the data down to summary statistics, and alter the BRL to use that new object, but that would be a whole different project!}
